---
title: "STA442 Assignment 2"
author: "Xin Wei"
date: "16/10/2019"
output:
  pdf_document: default
---

\section{Gap of Math Scores Can be Narrowed by Changing to Another School?}

In our analysis of the mathematics achievement scores between individuals from different schools, we found that students' math scores tend to vary a lot within each school, but the school-to-school variation does not weight much to the overall variation in math scores in general. 

As a basic subject in science, mathematics is complicated for many students. Nevertheless, some are still talented in it and get much enjoyment out of it. We are intereted in the topic that whether the varied performances in learning math is just a general phenomenon, or some school may have better teaching qualities, so that the gap among their students can be narrower.

We used the data frame "MathAchieve" in the MEMSS package. There are 7185 observations(individual students) and 160 groups(schools they attend) in our dataset. We want to explore which factor contributes more to the variation of math scores between different students. The two major factors(random effets) we focus on are the school-to-school variation and the individual variation. As an explanatory tool, we fit a general linear mixed model(which is hierarchical multi-level) as follows:

\begin{equation}
MathAch_{i}=\beta_{0} + X\beta + S_{i} + T_{ij}
\end{equation}

where X includes fixed effects Minority, Sex and SES(social-economic status). We treat the school($S_{i}$) one attends as a random effect, which measures the differences between the average scores at school i and in the entire sample dataset. And individual student($T_{ij}$) is another random effect. The restricted maximum likelihood estimation(REML) is conducted here, since compared with maximum likelihood estimates(ML), it gives relatively less biased random effect estimates. 

By looking at the table below, we found that the variation at level 1(school-to-school variation, $\sigma$) is around 1.92 and the variation at level 2(individual variation, $\tau$) is approximately 5.99. Calculating the variance ratio gives us $5.99^{2}$ / ($1.92^{2}$ + $5.99^{2}$) = 90.7%, which means the variation of individuals themselves contributed 90.7% to the total variation in the dataset. It is nearly as 10 times as the variation among different schools. In other words, the differences within schools are much larger than between students from different schools. This being said, we can conclude that the math teaching quality in different schools, such as the teaching methods, course arrangements, faculty, etc., may not vary a lot. But it is the math learning skills of students themselves that matters. One possible reason is that, it may be a common phenomenon in each school that, some students are more interested in this subject and put much effort into it, while others may not contribute much attention to their learning habits in math. 

```{r, echo=FALSE, message=FALSE}
library("nlme")
library("INLA")
data("MathAchieve", package = "MEMSS")
mod <- lme(MathAch ~ Minority + Sex + SES, random = ~1 | School, 
           data=MathAchieve)
tab1 <- Pmisc::lmeTable(mod)[,-3]
knitr::kable(tab1, digits = 2, escape=FALSE, 
             caption = "Linear Mixed Effect Table")
```

```{r, echo=FALSE, message=FALSE}
download.file("http://pbrown.ca/teaching/appliedstats/data/drugs.rds",
              "drugs.rds")
xSub = readRDS("drugs.rds")
```

\pagebreak

\section{Abusers of Cocaine in Virginia Less Likely to Rehabilitate?}
\subsection{Summary}
In the analysis of annual data of people discharged from substance abuse treatment programs, we found that young people addicted to alchohol and marijuana are more likely to complete their treatments, compared to those addicted to "hard" drugs, including Heroin, Opiates, Methamphetamine, and Cocaine. Apart from that, we also came up with evidence that some American states indeed have better treatment programs, and the completion rates in these states were much higher.

\subsection{Introduction}
As it is known that the abuse of substances(drugs, achohol, cocaine, etc.) causes serious health problems and affects the quality of lives, some abusers are actively involved in abuse treatment programs. We used the data provided by Treatment Episode Data Set -- Discharges (TEDS-D), trying to explore whether young people addicted to "hard" drugs(Heroin, Opiates, Methamphetamine, Cocaine) are less likely to complete treatment programs. We also want to know whether there are significant differences among American states in the efficiency of treatment programs. These are administrative records data, collected using longitudinal time method.

\subsection{Methods}
We used the dataset released by ICPSR. In the study during 2011, administrative records method was used to collect data from substance abuse treatment programs as reported to state substance abuse agencies. As an exploratory tool, based on historical data, we assumed penalized complexity prior distribution for our nested random effects, American states and towns. Specifically, for states, we assumed the probability of point estimates greater than 0.9 being 0.05, and for towns , we assumed P($\mu$ > 0.8) = 0.05. After obtaining the data, we used INLA to perform the approximate Bayesian inference for the posterior distribution, which is proportional to prior distribution. We used glmm method to fit a logistic model, since we are willing to know the relationship between the binary variable(completed the program or not) and a combination of fixed effects(substances, age, gender, ethnicity, homeless or not) and random effects(states, town). All main effects were retained, eliminating effects not significantly interacted with them. Our model is as follows:

\begin{equation}
\log(odds) = X\beta + S_{i} + T_{ij}
\end{equation}

\subsection{Results}
The odds of treatment completion table shows that holding the other effects constant, we can fit a logistic model with respect to the fixed effect, substances(alchohol, heroin, etc.). In this model, we treated marijuana as the reference group, the odds ratio between marijuana and alchohol is 1:1.642, which means people in the alchohol group were more likely to complete the treatment. Also, based on the data, there is a 95% probability that the true effect mean is in the interval [1.608, 1.677]. As for the other four groups(Heroin, Opiates, Methamphetamine, Cocaine), the odds are relatively lower than the odds of marijuana group. Hence, we have found some evidence that people addicted to "hard" drugs are less likely to complete their treatments, compared with marijuana, and alchohol as well. 
Then we want to analyze the differences in the efficiency of treatment programs among states in the U.S.. Indeed, we found that treatment programs in some states/towns performed much better than others. As shown in the table below, we found that the point estimate(posteior mean) of Virginia, -2.9, is the lowest of all states. And there is a 95% probability that the estimate falls within the credible interval [-3.2, -2.5]. Following are New Mexico(-1.1), North California(-0.8), etc. This indicates that programs in these states showed very low efficiency in treating people, and they negatively affected the average rate of program completion over the country. In contrast, it looks like some other states, such as Colorado, Florida, Massachusetts, etc., have much more efficient treatment programs and contributed positive effects to the overall average completion rate. Especially for Florida, its point estimate, 1.0, is significantly 3.9 higher than that of Virginia, and the credible interval [0.7, 1.3] confirms that with a probablity of 95%, the estimate would be at least 0.7.



```{r, echo=FALSE, message=FALSE}
forInla = na.omit(xSub)
forInla$y = as.numeric(forInla$completed)
library("raster")
library("INLA")
library("Pmisc")
library("Hmisc")
library("data.table")
ires = inla(y ~ SUB1 + GENDER + raceEthnicity + homeless + 
              f(STFIPS,hyper=list(prec=list(
                prior='pc.prec', param=c(0.9, 0.05)))) + 
              f(TOWN, hyper=list(prec=list(
                prior='pc.prec', param=c(0.8, 0.05)))), 
            data=forInla, 
            family='binomial', 
            inla.link.logit,
            control.inla = list(strategy='gaussian',
                                int.strategy='eb'))
            
sdState = Pmisc::priorPostSd(ires)

do.call(matplot, sdState$STFIPS$matplot)
do.call(legend, sdState$legend)

do.call(matplot, sdState$TOWN$matplot)
do.call(legend, sdState$legend)


```


\pagebreak

```{r, echo=FALSE, message=FALSE}
toPrint = as.data.frame(rbind(exp(ires$summary.fixed[, c(4, 3, 5)]),
                              sdState$summary[, c(4, 3, 5)]))
sss = "^(raceEthnicity|SUB1|GENDER|homeless|SD)(.[[:digit:]]+.[[:space:]]+| for )?"
toPrint = cbind(variable = gsub(paste0(sss, ".*"),
                                "\\1", 
                                rownames(toPrint)), 
                category = substr(gsub(sss, "", 
                                       rownames(toPrint)), 1, 25),
                toPrint)
Pmisc::mdTable(toPrint, digits = 3, mdToTex = TRUE, 
               guessGroup = TRUE, 
               caption = "Posterior Means and Credible Intervals 
               for Model Effects")

```


```{r, echo=FALSE, message=FALSE}
ires$summary.random$STFIPS$ID = gsub("[[:punct:]]|[[:digit:]]",
                                     "", ires$summary.random$STFIPS$ID)
ires$summary.random$STFIPS$ID = gsub("DISTRICT OF COLUMBIA",
                                     "WASHINGTON DC", ires$summary.random$STFIPS$ID)
toprint = cbind(ires$summary.random$STFIPS[1:26, c(1, 2, 4, 6)],
                ires$summary.random$STFIPS[-(1:26),
                                           c(1, 2, 4, 6)])
colnames(toprint) = gsub("uant", "", colnames(toprint))
knitr::kable(toprint, digits = 1, 
             caption = "Posterior Means and Credible 
             Intervals for States")
```

\pagebreak
\subsection{Appendix}
```{r, message=FALSE, eval=FALSE}
library("nlme")
library("INLA")
data("MathAchieve", package = "MEMSS")
mod <- lme(MathAch ~ Minority + Sex + SES, random = ~1 | School, 
           data=MathAchieve)
tab1 <- Pmisc::lmeTable(mod)[,-3]
knitr::kable(tab1, digits = 2, escape=FALSE, 
             caption = "Linear Mixed Effect Table")

download.file("http://pbrown.ca/teaching/appliedstats/data/drugs.rds",
              "drugs.rds")
xSub = readRDS("drugs.rds")

forInla = na.omit(xSub)
forInla$y = as.numeric(forInla$completed)
library("raster")
library("INLA")
library("Pmisc")
library("Hmisc")
library("data.table")
ires = inla(y ~ SUB1 + GENDER + raceEthnicity + homeless + 
              f(STFIPS,hyper=list(prec=list(
                prior='pc.prec', param=c(0.9, 0.05)))) + 
              f(TOWN, hyper=list(prec=list(
                prior='pc.prec', param=c(0.8, 0.05)))), 
            data=forInla, 
            family='binomial', 
            inla.link.logit,
            control.inla = list(strategy='gaussian',
                                int.strategy='eb'))
            
sdState = Pmisc::priorPostSd(ires)

do.call(matplot, sdState$STFIPS$matplot)
do.call(legend, sdState$legend)

do.call(matplot, sdState$TOWN$matplot)
do.call(legend, sdState$legend)

toPrint = as.data.frame(rbind(exp(ires$summary.fixed[, c(4, 3, 5)]),
                              sdState$summary[, c(4, 3, 5)]))
sss = "^(raceEthnicity|SUB1|GENDER|homeless|SD)
(.[[:digit:]]+.[[:space:]]+| for )?"
toPrint = cbind(variable = gsub(paste0(sss, ".*"),
                                "\\1", 
                                rownames(toPrint)), 
                category = substr(gsub(sss, "", 
                                       rownames(toPrint)), 1, 25),
                toPrint)

Pmisc::mdTable(toPrint, digits = 3, mdToTex = TRUE, 
               guessGroup = TRUE, 
               caption = "Posterior Means and Credible Intervals 
               for Model Effects")
               
ires$summary.random$STFIPS$ID = gsub("[[:punct:]]|[[:digit:]]",
                                     "", ires$summary.random$STFIPS$ID)
ires$summary.random$STFIPS$ID = gsub("DISTRICT OF COLUMBIA",
                                     "WASHINGTON DC", ires$summary.random$STFIPS$ID)
toprint = cbind(ires$summary.random$STFIPS[1:26, c(1, 2, 4, 6)],
                ires$summary.random$STFIPS[-(1:26),
                                           c(1, 2, 4, 6)])
colnames(toprint) = gsub("uant", "", colnames(toprint))
knitr::kable(toprint, digits = 1, 
             caption = "Posterior Means and Credible 
             Intervals for States")

```
